The Key to a Data Parallel Compiler

### 2016-04-07 2016-spring-array/2016-array-key.tex
Cleanup the main sections

That is, the sample code given indented by 6 spaces, followed by the value 
of the expression without indent. All expressions are evaluated right to 
left and all function application is unified and infix; that is, a function 
application may apply a function (such as \verb;+;) to one or two arguments, 
which must appear on the right and (optionally) on the left of the function 
name, as in the above example. Tables \ref{tab:scalarprims} to 
\ref{tab:conjunctions}
provide a listing of all of the primitives used in the Co-dfns compiler, 
a selection of which are used here.

We separate out higher-order functions called operators, from functions that 
operate over arrays that we call simply functions. Operators take a function 
argument or two, and bind more strongly to the left than the right. An 
operator may take one or two operands, either on the left, or on the 
right and left, and will return a function. In the following 
example, we apply a reduction operator to a function created using the 
composition operator, as an example:

\begin{verbatim}
      +∘-/1 2 3 4 5
3
      +/1 2 3 4 5
15
\end{verbatim}

Uses of the notation are introduced slowly throughout the treatment, so 
the reader is encouraged to simply proceed through the treatment, referencing 
the tables only when required.

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic            & Dyadic \\
\midrule
\texttt{+}               & Identity           & Plus (Add) \\
\texttt{-}               & Negative           & Minus (Subtract) \\
\texttt{×}               & Direction (Signum) & Times (Multiply) \\
\texttt{÷}               & Reciprocal         & Divide \\
\texttt{|}               & Magnitude          & Residue (Modulo) \\
\texttt{⌊}               & Floor              & Minimum \\
\texttt{⌈}               & Ceiling            & Maximum \\
\texttt{*}               & Exponential        & Power \\
\texttt{⍟}               & Natural Logarithm  & Logarithm \\
\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
\texttt{!}               & Factorial          & Binomial \\
\texttt{\textasciitilde} & Not                & \\
\texttt{?}               & Roll               & \\
\texttt{∧}              &                    & And \\
\texttt{∨}               &                    & Or \\
\texttt{⍲}               &                    & Nand \\
\texttt{⍱}               &                    & Nor \\
\texttt{<}               &                    & Less \\
\texttt{≤}               &                    & Less Or Equal \\
\texttt{=}               &                    & Equal \\
\texttt{≥}               &                    & Greater Or Equal \\
\texttt{>}               &                    & Greater \\
\texttt{≠}               &                    & Not Equal \\
\end{tabular}
\caption{Scalar Primitives}
\label{tab:scalarprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic       & Dyadic \\
\midrule
\texttt{⍴} & Reshape       & \\
\texttt{,} & Ravel         & Catenate/Laminate \\
\texttt{⍪} & Table         & Catenate First/Laminate \\
\texttt{⌽} & Reverse       & Rotate \\
\texttt{⊖} & Reverse First & Rotate First \\
\texttt{⍉} & Transpose     & Transpose \\
\texttt{↑} & Mix           & \\
\texttt{↓} & Split         & \\
\texttt{⊂} & Enclose       & Partitioned Enclose \\
\texttt{∊} & Enlist        & \\
\end{tabular}
\caption{Structural Mixed Primitives}
\label{tab:structprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic  & Dyadic \\
\midrule
\texttt{⊃}               & Disclose & Pick\\
\texttt{↑}               &          & Take \\
\texttt{↓}               &          & Drop \\
\texttt{/}               &          & Replicate \\
\texttt{⌿}               &          & Replicate First \\
\texttt{\textbackslash}  &          & Expand \\
\texttt{⍀}               &          & Expand First \\
\texttt{\textasciitilde} &          & Without (Excluding) \\
\texttt{∩}               &          & Intersection \\
\texttt{∪}               & Unique   & Union \\
\texttt{⊣}               & Same     & Left \\
\texttt{⊢}               & Identity & Right\\
\end{tabular}
\caption{Selection Mixed Primitives}
\label{tab:selprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic         & Dyadic \\
\midrule
\texttt{⍳} & Index Generator & Index Of \\
\texttt{∊} &                 & Membership \\
\texttt{⍋} & Grade Up        & Grade Up \\
\texttt{⍒} & Grade Down      & Grade Down \\
\texttt{?} &                 & Deal \\
\texttt{⍷} &                 & Find \\
\end{tabular}
\caption{Selector Mixed Primitives}
\label{tab:seltorprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic       & Dyadic \\
\midrule
\texttt{⍴} & Shape         & \\
\texttt{≡} & Depth         & Match \\
\texttt{≢} & Tally         & Not Match \\
\texttt{⍎} & Execute       & Execute \\
\texttt{⍕} & Format        & Format \\
\texttt{⊥} &               & Decode (Base) \\
\texttt{⊤} &               & Encode (Representation) \\
\texttt{⌹} & Matrix Divide & Matrix Inverse \\
\end{tabular}
\caption{Miscellaneous Mixed Primitives}
\label{tab:miscprims}
\end{table}

@@

We divide the work of sub-tree computation into three basic phases. Firstly, 
we maintain a node coordinate with each node that uniquely identifies it 
relative to others, we use these coordinates to select and group nodes for 
work as a single sub-tree, and then we operate over these groupings using 
the key operator.

@@

We represent the AST as a matrix of 3 columns and n rows, one for
each node in the tree. The first column contains the inter-node
relationships in the form of a depth vector. The second column is a
vector of the node types, while the third contains the ``value''
of the node, such as the name in a variable.

Parsing the examples above gives the following ASTs:

@@

The depth vector for these trees we name \verb;Fd; and \verb;Ed;, respectively:

\begin{verbatim}
Fd←0 1 2 3 4 4 4 2 2 3
Ed←0 1 2 2 2 1 1 2 2 2
\end{verbatim}

@@

\begin{verbatim}
Ft←’FEFEVPVAN’
Et←’EEVPVPEVPV’
\end{verbatim}

@@

\begin{verbatim}
Fv←’f’ 0 0 0 ‘⍵’ ‘+’ 0 7
Ev←’v’ 0 ‘a’ ‘+’ ‘b’ ‘÷’ 0 ‘c’ ‘×’ ‘d’
\end{verbatim}

We combine these to form the respective AST matrices. We write \verb;A,B;
to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a
vector into a 1-column matrix. Thus, the two tree matrices are given
by the following expressions:

@@

Note especially that all inter-node information lives in the depth
vector, but that this information requires non-local access as
described in the previous section to use it. Constructing node
coordinates fixes this issue.

@@

Every node in an AST has a node coordinate, named because a coordinate
is a precise location in a space.  We can imagine all the nodes
arranged inside some multi-dimensional space, leading to a specific
set of node coordinates. Many such arrangements exist, of varying
usefulness. In our case, any arrangement should allow us to answer
the following questions about any two arbitrary nodes in a tree:

@@

In short, we care about the relative position of each node in a tree
relative to any other. We define a node coordinate as follows.

@@

To understand how we construct these coordinates we must consider
how we represent the AST.

@@

In building the node coordinates, our goal is to build a matrix where
each row is a node coordinate satisfying our previously described
requirements. We end up with an n d shaped matrix where n is the number
of rows and d the depth of the tree. As noted above, the depth vector
contains all the necessary information, but in the wrong form.

We write \verb;f⌿A; to reduce the first axis of \verb;A; using function \verb;f;. Thus,
\verb;+⌿V; is the sum of the elements in vector \verb;V;. The function \verb;x⌈y;
gives the maximum of its two arguments. So, the depth of each tree
is given by the following:

@@

We can obtain the ordered sequence  by writing \verb;⍳n;:

@@

So the depths of all nodes that appear in the depth vectors is thus:

@@

The function table or outer product of \verb;f; over vectors \verb;U; and \verb;V; is
written \verb;U ∘.f V; giving a \verb;U V; shaped matrix as a result. Thus,
\verb;(⍳3)∘.×⍳3; gives a small multiplication table:

@@

If we use \verb;=; instead, we have a Boolean identity matrix:

@@

If we use \verb;∘.=; on the depth vector and its set of depths instead,
we see an expanded Boolean representation of the depth vector:

@@

These matrices let us see the nesting features of each tree more
visually, but also suggest another step. We can compute a sum scan
with \verb;+⍀;, also called a prefix sum, along the first axis. Applying
this function on the above matrices leads to an interesting result:

@@

These matrices are lexicographically ordered, and each ancestor
shares a common prefix with its descendants. They are also unique
coordinates. Only the spurious digits at the end of each coordinate
prevent these matrices from meeting all our requirements for valid
node coordinates.

@@

If \verb;n↑V; takes the first \verb;n; elements of \verb;V;, then we can obtain coordinate
matrices from the prefix sums by noting that the spurious digits
all come after column \verb;d+1; where \verb;d; is the depth corresponding to that
coordinate. The following gives a complete expression for computing
a node coordinate matrix from a depth vector, shown using \verb;Fd; and \verb;Fe;:

@@

The simplest operation over a node coordinate is to extract the depth
of the node. The expression \verb;C⍳0; finds the first occurrence of \verb;0; in \verb;C;
and returns the index of that occurrence. (The application of ⍳ here
is its dyadic form, meaning that it receives two arguments. Previous
uses of ⍳ have used its monadic form, which generates all indices,
instead of finding a specific index. Most APL functions have a
dyadic and monadic form.)  Thus, the depth of a node is given by
the following:

@@

The primary calculation we care about in order to do function lifting,
expression flattening, and other sorts of flattening transformations
is to determine which nodes are ancestors of another node. The core
of this computation determines whether one node is a child of another
based on their node coordinates. This amounts to computing whether
one node is a prefix of another, ignoring zeros.

We write \verb;(f g h); to represent the composition of functions \verb;f;, \verb;g;,
and \verb;h; as a function train. This is used with another shorthand of
composition according to the following equivalences:

@@

With this, we can write a function to compute whether a given
coordinate is a prefix of another.

@@

Here \verb;P; is the coordinate we wish to check against \verb;C;, to determine
whether \verb;C; is a descendant. The functions are all standard logical
functions extended to apply point-wise over arrays. The function
\verb;⊢; always returns its right argument. The function \verb;(=∨0=⊢);
reads intuitively as, ``equal or right element is zero.'' In this
case to make this a predicate we combine this with the ``for all''
reduction using \verb;∧⌿;. This particular pattern is a special case
of inner product, which we can compute using \verb;f.g;. Thus, \verb;+.×; is a
function that computes traditional matrix-matrix multiplication when
applied to two matrices. However, we can also use it to compute the
above reduction and application:

\begin{verbatim}
      C∧.(=∨0=⊢)P
1
\end{verbatim}

This extends the reduction to allow us to use matrix values for
\verb;C; and \verb;P; instead of vectors, and thus determine whether rows in \verb;C;
are descendants of multiple candidates given by \verb;P;. This becomes
important to using the prefix function. In our examples, we want
to determine the nearest ancestor of a certain node type to which a
node belongs. In the case of function lifting, we want to determine
the nearest ancestor function, in the case of expression flattening,
we want to determine the nearest ancestor expression node. We can
extract the rows we care about for each example using \verb;⌷;, where \verb;i⌷M;
returns the i-th row of \verb;M;. In the following expressions, we use \verb;0 2;
rather than \verb;¯1; in \verb;⍤; to indicate that \verb;⌷; will receive the entire
matrix on the right for each scalar index on the left.

\begin{verbatim}
      ⊢Fp←0 2⌷⍤0 2⊢Fc
1 0 0 0 0
1 1 1 0 0
      ⊢Ep←0 1 6⌷⍤0 2⊢Ec
1 0 0
1 1 0
1 3 0
\end{verbatim}

The matrices \verb;Fp; and \verb;Ep; correspond to the function and expression
node coordinates in our two examples, respectively. We could now
use \verb;Fp; and \verb;Ep; to compare against \verb;Fc; and \verb;Ec; and determine which nodes
contained which nodes in the tree. However, our prefix function will
return 1 when we ask whether a node is a prefix of itself. In this
case we don’t want that. The solution to this is to drop the last
non-zero element from the coordinate. This will not prevent it from
matching against any of its ancestors, but will prevent matching
against itself. We can use the depth vector with the \verb;↑; function to
take all but the last non-zero element. We need to remember to extend
the returned array with an extra zero column, since the shape will
be too small otherwise.

@@

We can now use \verb;Fcp; and \verb;Ecp; to determine which Function and Expression
nodes match for each node:

@@

We use the ⍉ function, which computes the transpose of its right
argument, to ensure that the row and column sizes match up. With the
above, we can determine the ``greatest'' match, which is the closest.
We can replace each 1 in the above matrices by their column numbers
(the number of the column in which that 1 occurs), and then we can
take the maximum column number in each row to determine an index in
either \verb;Fp; or \verb;Ep; that is the appropriate ``parent'' coordinate for
that node. This works because we have made sure to order the \verb;Ep; and \verb;Fp;
matrices lexicographically. The following shows this computation for
the expression example; we elide the function example as redundant
at this point. We also use \verb;⌈/; instead of \verb;⌈⌿; to reduce along
the last, rather than the first, axis.

\begin{verbatim}
      ⊢Ei←⌈/(⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
0 0 1 1 1 0 0 2 2 2
      ⊢Ek←Ei⌷⍤0 2⊢Ep 
1 0 0
1 0 0
1 1 0
1 1 0
1 1 0
1 0 0
1 0 0
1 3 0
1 3 0
1 3 0
\end{verbatim}

At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the
closest containing node that we care about for each node in the tree,
using its node coordinate.

\begin{verbatim}
      Fk
1 0 0 0 0
1 0 0 0 0
1 0 0 0 0
1 1 1 0 0
1 1 1 0 0
1 1 1 0 0
1 1 1 0 0
1 0 0 0 0
1 0 0 0 0
\end{verbatim}

We will use these keys to perform computation over the AST and
accomplish our tasks of function lifting and expression flattening in
the next section, but we make a final note here, that we can imagine
many other operations which might be used throughout the compiler when
we care about how nodes relate to each other. The use of operators
like reduction, scan, inner and outer products allow us to obtain
the necessary information from the node coordinate matrix. Sometimes
these results might not be obvious, but the example of the ``belongs
to'' relationship given above demonstrates a pattern that arises
repeatedly throughout our data parallel compiler, and is therefore
particularly useful to understand.

@@

The Key operator (written \verb;⌸;), takes a single function
that expects two arguments and returns a function which takes two
arguments. The left argument is a set of keys, and the right argument
is a set of corresponding elements associated with those keys. In our
case, we provide matrices for both of these arguments, so each pair
of rows corresponds to a key-value pair. A simple example of the key
operation at work is to compute a histogram (\verb;≢; here can be thought
of as ``tally'' or count):

\begin{verbatim}
       ⊢X←?10⍴5
1 1 4 0 1 3 1 1 2 1
      X(⊣,(≢⊢))⌸X
1 6
4 1
0 1
3 1
2 1
\end{verbatim}

@@

The following compiler passes are often found in any suitably complex 
language being compiled, and represent a difficult problem for anyone attempting 
to do data-parallel computation. Indeed, they were the primary motivating 
factor for the development of the node coordinate matrix and its associated 
techniques. They demonstrate how we can operate on highly nested structures, 
usually with the intent of eliminating that structure to more closely map 
the intended output language.

These case studies come from the Co-dfns compiler, which is compiling a 
lexically scoped language that contains both nested functions and 
expressions.

@@

If we use the key operator with \verb;Fk;, we get the following (the monadic
use of \verb;↓; converts \verb;Fc; from a matrix to a vector of vectors):

@@

@@

The J programming language \cite{hui2014key} was the first practical,
general-purpose programming language to introduce the key operator as a
primitive operator with the presumption of its general usefulness. The
rank operator (⍤) used throughout this treatment also derives from
the J traditional, receiving particular interest throughout the APL
community \cite{bernecky1987rank,hui1995rank}.

@@

>>>>

That is, the sample code given indented by 6 spaces, followed by the value 
of the expression without indent. All expressions are evaluated right to 
left and all function application is unified and infix; that is, a function 
application may apply a function (such as \verb;+;) to one or two arguments, 
called monadic and dyadic application respectively,
which must appear on the right and (optionally) on the left of the function 
name, as in the above example. Tables \ref{tab:scalarprims} to 
\ref{tab:conjunctions}
provide a listing of all of the primitives used in the Co-dfns compiler, 
a selection of which are used here.

We separate out higher-order functions called operators, from functions that 
operate over arrays that we call simply functions. Operators take a function 
argument or two, and bind more strongly to the left than the right. An 
operator may take one or two operands, either on the left, or on the 
right and left, and will return a function. In the following 
example, we apply the reduction operator (\verb;/;) to a function created using the 
composition operator (\verb;∘;), as an example:

\begin{verbatim}
      +/1 2 3 4 5
15
      1+-(2+-(3+-(4+-(5))))
3
      1 +∘- 2 +∘- 3 +∘- 4 +∘- 5
3
      +∘-/1 2 3 4 5
3
\end{verbatim}

We introduce notation as needed throughout the exposition, so the reader 
is encouraged to refer to the tables as necessary to recall particular 
operations.

@@

We divide the work of sub-tree computation into three basic phases:  
we maintain a node coordinate with each node that uniquely identifies it 
relative to others, we use these coordinates to select and group nodes for 
work as a single sub-tree, and then we operate over these groupings using 
the Key operator (\verb;⌸;).

@@

We represent the AST as a 3 column matrix with one row per node in the tree.
The first column contains the inter-node
relationships in the form of a depth vector. The second column is a
vector of the node types, while the third contains the ``value''
of the node, such as the name in a variable.

We will use the following two running examples throughout. The \verb;F; 
tree is an example nested function, while the \verb;E; tree is an example 
nested expression.

@@

The depth vectors for these trees we name \verb;Fd; and \verb;Ed;, respectively:

\begin{verbatim}
Fd←0 1 2 3 4 4 4 2 3
Ed←0 1 2 2 2 1 1 2 2 2
\end{verbatim}

@@

\begin{verbatim}
Ft←'FEFEVPVAN'
Et←'EEVPVPEVPV'
\end{verbatim}

@@

\begin{verbatim}
Fv←'f' 0 0 0 '⍵' '+' '⍵' 0 7
Ev←'v' 0 'a' '+' 'b' '÷' 0 'c' '×' 'd'
\end{verbatim}

We combine these to form the respective AST matrices. We write \verb;A,B;
to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a
vector into a 1-column matrix. Thus, the two AST matrices are given
by the following expressions:

@@

The Depth Vector stores all edge information for the tree, but this information 
requires non-local access to utilize, such as traversing potentially the 
entire vector to determine the children of a specific node.
Node Coordinates fix this issue.

@@

Every node in an AST has a Node Coordinate, named because a coordinate
is a precise location in a space.  We can imagine all the nodes
arranged inside some multi-dimensional space, leading to a specific
set of node coordinates. Many such arrangements exist, of varying
usefulness. In our case, any arrangement should allow us to answer
the following questions about any two arbitrary nodes in a tree:

@@

\noindent
In short, we care about the relative position of each node in a tree
relative to any other. We define a node coordinate as follows.

@@

@@

We construct the Node Coordinate Matrix from the depth vector, thereby encoding 
the depth vector into a more useful format. The matrix has \verb;N; rows and 
\verb;D; columns, where \verb;N; is the node count and \verb;D; is the depth 
of the tree.

We write \verb;f⌿A; to reduce the first axis of \verb;A; using function \verb;f;. Thus,
\verb;+⌿V; is the sum of the elements in vector \verb;V;. The function \verb;x⌈y;
gives the maximum of its two arguments.
We compute the depth of each tree as follows:

@@

\noindent
We can obtain the ordered sequence  by writing \verb;⍳n;:

@@

\noindent
So the depths of all nodes that appear in the depth vectors is thus:

@@

\noindent
The function table or outer product of \verb;f; over vectors \verb;U; and \verb;V; is
written \verb;U ∘.f V; giving a \verb;U V; shaped matrix as a result. Thus,
\verb;(⍳3)∘.×⍳3; gives a small multiplication table:

@@

\noindent
If we use \verb;=; instead, we have a Boolean identity matrix:

@@

\noindent
If we use \verb;∘.=; on the depth vector and its set of depths instead,
we see an expanded Boolean representation of the depth vector:

@@

\noindent
These matrices let us see the nesting features of each tree more
visually, but also suggest another step. We can compute a sum scan
with \verb;+⍀;, also called a prefix sum, along the first axis. Applying
this function on the above matrices leads to an interesting result:

@@

\noindent
These matrices are lexicographically ordered, and each ancestor
shares a common prefix with its descendants. They are also unique
coordinates. Only the spurious digits at the end of each coordinate
prevent these matrices from meeting all our requirements for valid
node coordinates.

@@

\noindent
If \verb;n↑V; takes the first \verb;n; elements of \verb;V;, then we can obtain coordinate
matrices from the prefix sums by noting that the spurious digits
all come after column \verb;d+1; where \verb;d; is the depth corresponding to that
coordinate. The following gives a complete expression for computing
a node coordinate matrix from a depth vector, shown using \verb;Fd; and \verb;Fe;:

@@

The simplest operation over a node coordinate is to extract the depth
of the node.
The dyadic expression \verb;C⍳0; finds the first occurrence of \verb;0;
in \verb;C; and returns the index of that occurrence.
Thus, we compute the depth of a node as follows:

@@

\noindent
In many compiler passes, we primarily care about which nodes are ancestors 
of other nodes. To determine whether one node is a child of another, we compute 
whether one node is a prefix of the other, ignoring zeros.
We write \verb;(f g h); to represent the composition of functions \verb;f;, \verb;g;,
and \verb;h; as a function train.
Function trains obey the following equivalences, where \verb;A; and \verb;B;
are arrays.

@@

\noindent
With this, we write a function to compute whether a given coordinate prefixes 
another.

@@

\noindent
The above determines whether \verb;P; is an ancestor to \verb;C;.
The logical functions \verb;=;, \verb;∧;, and \verb;∨; are all extended 
point-wise over arrays.
The function \verb;⊢; returns its right argument.
The function \verb;(=∨0=⊢); reads as, ``equal or a zero right argument.'' 
We take this point-wise operation and reduce it with \verb;∧⌿; (For All).
This pattern is a special case of inner product, written \verb;f.g;. 
For example, \verb;+.×; is matrix multiplication. We transform our reduction and 
boolean function to a single predicate with the inner product operator:
\begin{verbatim}
      C∧.(=∨0=⊢)P
1
\end{verbatim}

The use of inner product extends to cases where \verb;C; or \verb;P; are matrices.
For our examples, we want to lift the functions and flatten expressions, so we 
will select the \verb;F; nodes and \verb;E; nodes, respectively, as our 
parent/ancestor nodes in \verb;Fp; and \verb;Ep;. Given a boolean vector 
\verb;BV;, the expression \verb;BV⌿M; selects rows from \verb;M; based on the 
non-zero elements of \verb;BV;.

\begin{verbatim}
      ⊢Fp←('F'=Ft)⌿Fc
1 0 0 0 0
1 1 1 0 0
      ⊢Ep←('E'=Et)⌿Ec
1 0 0
1 1 0
1 3 0
\end{verbatim}

\noindent
We could use \verb;Fp; and \verb;Ep; to compare against \verb;Fc; and \verb;Ec;
to determine which nodes belong to which parent in \verb;Fp; or \verb;Ep;, 
but our prefix function returns 1 when \verb;C≡P;. 
Instead, we drop the last non-zero element from each coordinate.
This will permit matching against all ancestors, but not against itself.
We use the depth vector and the Take (\verb;↑;) function to take all but
the last non-zero element. We extend the resulting array with an extra 
zero column to ensure we have a compatible shape.

@@

\noindent
We use \verb;Fcp; and \verb;Ecp; to determine the \verb;F; and \verb;E; 
ancestors for each node (\verb;⍉A; transposes \verb;A;):

@@

\noindent
In the above, the closest ancestor is the rightmost 1 in each row.
We can extract the column number of the rightmost 1 by replacing each 1 
with its column number and selecting the maximum of each row with \verb;⌈/;. 
For space we will show only the \verb;E; example.
In the following, \verb;V×⍤1⊢M; gives a matrix where each row is given by 
\verb;V×i⌷M; where \verb;i⌷M; is the $i$th row of \verb;M;.

\begin{verbatim}
      (⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
0 0 0
0 0 0
0 1 0
0 1 0
0 1 0
0 0 0
0 0 0
0 0 2
0 0 2
0 0 2
      ⊢Ei←⌈/(⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
0 0 1 1 1 0 0 2 2 2
\end{verbatim}

\noindent
We use these column numbers to index into \verb;Ep; to obtain an ancestor 
matrix where the node coordinate of the closest ancestor is given for 
each node, one per row. 
The expression \verb;I⌷⍤0 2⊢M; gives a matrix, one row per index in \verb;I;, 
consisting of rows of \verb;M; selected by row indices in \verb;I;.

\begin{verbatim}
      ⊢Ek←Ei⌷⍤0 2⊢Ep 
1 0 0
1 0 0
1 1 0
1 1 0
1 1 0
1 0 0
1 0 0
1 3 0
1 3 0
1 3 0
\end{verbatim}

\noindent
At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the
closest containing node that we care about for each node in the tree,
using its node coordinate.

\begin{verbatim}
      Fi←⌈/(⍳2)×⍤1⊢Fcp∧.(=∨0=⊢)⍉Fp
      ⊢Fk←Fi⌷⍤0 2⊢Fp
1 0 0 0 0
1 0 0 0 0
1 0 0 0 0
1 1 1 0 0
1 1 1 0 0
1 1 1 0 0
1 1 1 0 0
1 0 0 0 0
1 0 0 0 0
\end{verbatim}

\noindent
We use these keys to compute over the AST, particularly for function lifting 
and expression flattening as demonstrated in the next section. The repeating 
pattern is to leverage node coordinates and array operations to do stackless 
reasoning about inter-node relationships. This pattern occurs often in our 
compiler and is particularly useful to enable straightforward data-parallel 
transformations over an AST.

@@

The Key operater (written \verb;⌸;) is central to our strategy for compilation.
It allows us to use the Node Coordinates to their full effect. 
The expression \verb;K f⌸ M; groups rows of \verb;M; by their corresponding 
keys in \verb;K; and computes \verb;k f m; for each unique key \verb;k; in 
\verb;K; and the matrix \verb;m; of rows of \verb;M; with key \verb;k;. 
The corresponding rows of matrices \verb;K; and \verb;M; form key-value pairs.
A simple example of this at work is to compute a histogram using Tally (\verb;≢;):

\begin{verbatim}
      10⍴5
5 5 5 5 5 5 5 5 5 5
       ⊢X←?10⍴5
1 1 4 0 1 3 1 1 2 1
      ≢⍳5
5
      X(⊣,(≢⊢))⌸X
1 6
4 1
0 1
3 1
2 1
\end{verbatim}

@@

The following compiler passes are found in any compiler from a nested function, 
nested expression language to a language without nested functions 
or expressions. They were the chief motivating examples for developing the 
techniques presented above as a part of developing the Co-dfns compiler, 
and represent a challenging problem to data-parallel compilation without 
the above techniques, but fall out almost effortlessly once the above patterns 
are available.

@@

If we use the Key operator with \verb;Fk;, we get the following (the monadic
use of \verb;↓; converts \verb;Fc; from a matrix to a vector of vectors):

@@

This function lifting works even with lexically scoped functions, 
as is the case for the Co-dfns compiler.
We can do this because the node coordinates maintain the lexical information 
for each variable reference, enabling us to construct the appropriate lexical 
binding for each variable much later than would normally be done in a 
compiler. 
In other words, the above lifting operation loses no information that would 
prevent us from handling lexical scoping, but decouples lexical scoping 
from the act of lifting functions.

@@

The J programming language \cite{hui2014key} was the first practical,
general-purpose programming language to introduce the Key operator as a
primitive operator with the presumption of its general usefulness. The
rank operator (⍤) used throughout this treatment also derives from
the J traditional, receiving particular interest throughout the APL
community \cite{bernecky1987rank,hui1995rank}.

@@

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic            & Dyadic \\
\midrule
\texttt{+}               & Identity           & Plus (Add) \\
\texttt{-}               & Negative           & Minus (Subtract) \\
\texttt{×}               & Direction (Signum) & Times (Multiply) \\
\texttt{÷}               & Reciprocal         & Divide \\
\texttt{|}               & Magnitude          & Residue (Modulo) \\
\texttt{⌊}               & Floor              & Minimum \\
\texttt{⌈}               & Ceiling            & Maximum \\
\texttt{*}               & Exponential        & Power \\
\texttt{⍟}               & Natural Logarithm  & Logarithm \\
\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
\texttt{!}               & Factorial          & Binomial \\
\texttt{\textasciitilde} & Not                & \\
\texttt{?}               & Roll               & \\
\texttt{∧}              &                    & And \\
\texttt{∨}               &                    & Or \\
\texttt{⍲}               &                    & Nand \\
\texttt{⍱}               &                    & Nor \\
\texttt{<}               &                    & Less \\
\texttt{≤}               &                    & Less Or Equal \\
\texttt{=}               &                    & Equal \\
\texttt{≥}               &                    & Greater Or Equal \\
\texttt{>}               &                    & Greater \\
\texttt{≠}               &                    & Not Equal \\
\end{tabular}
\caption{Scalar Functions}
\label{tab:scalarprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic       & Dyadic \\
\midrule
\texttt{⍴} & Reshape       & \\
\texttt{,} & Ravel         & Catenate/Laminate \\
\texttt{⍪} & Table         & Catenate First/Laminate \\
\texttt{⌽} & Reverse       & Rotate \\
\texttt{⊖} & Reverse First & Rotate First \\
\texttt{⍉} & Transpose     & Transpose \\
\texttt{↑} & Mix           & \\
\texttt{↓} & Split         & \\
\texttt{⊂} & Enclose       & Partitioned Enclose \\
\texttt{∊} & Enlist        & \\
\end{tabular}
\caption{Structural Mixed Functions}
\label{tab:structprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic  & Dyadic \\
\midrule
\texttt{⊃}               & Disclose & Pick\\
\texttt{↑}               &          & Take \\
\texttt{↓}               &          & Drop \\
\texttt{/}               &          & Replicate \\
\texttt{⌿}               &          & Replicate First \\
\texttt{\textbackslash}  &          & Expand \\
\texttt{⍀}               &          & Expand First \\
\texttt{\textasciitilde} &          & Without (Excluding) \\
\texttt{∩}               &          & Intersection \\
\texttt{∪}               & Unique   & Union \\
\texttt{⊣}               & Same     & Left \\
\texttt{⊢}               & Identity & Right\\
\end{tabular}
\caption{Selection Mixed Functions}
\label{tab:selprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic         & Dyadic \\
\midrule
\texttt{⍳} & Index Generator & Index Of \\
\texttt{∊} &                 & Membership \\
\texttt{⍋} & Grade Up        & Grade Up \\
\texttt{⍒} & Grade Down      & Grade Down \\
\texttt{?} &                 & Deal \\
\texttt{⍷} &                 & Find \\
\end{tabular}
\caption{Selector Mixed Functions}
\label{tab:seltorprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic       & Dyadic \\
\midrule
\texttt{⍴} & Shape         & \\
\texttt{≡} & Depth         & Match \\
\texttt{≢} & Tally         & Not Match \\
\texttt{⍎} & Execute       & Execute \\
\texttt{⍕} & Format        & Format \\
\texttt{⊥} &               & Decode (Base) \\
\texttt{⊤} &               & Encode (Representation) \\
\texttt{⌹} & Matrix Divide & Matrix Inverse \\
\end{tabular}
\caption{Miscellaneous Mixed Functions}
\label{tab:miscprims}
\end{table}
